---
title: "Data Cleaning & EDA"
output: html_document
date: "2024-11-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# remove objects
rm(list = ls())

# Library packages
library(tidyverse)
library(dplyr)
library(kableExtra)
library(psych)
library(GGally)

```

# Data Cleaning

```{r}
# Read the data
sales.zipcodes <- read.csv("project.sales.zipcodes.csv")
sales.cities <- read.csv("project.sales.cities.csv")
sales.counties <- read.csv("project.sales.counties.csv")

acs.zipcodes <- read.csv("project.acs.zipcodes.csv")
acs.cities <- read.csv("project.acs.cities.csv")
acs.counties <- read.csv("project.acs.counties.csv")
```

```{r}
# Merge data for country, city and zipcode levels
counties_data <- sales.counties %>%
  left_join(acs.counties, by = "county") %>%
  # Convert 'county' and 'category' to factors
  mutate(across(c(county, category), as.factor))
  
cities_data <- sales.cities %>%
  left_join(acs.cities, by = "city") %>%
  # Convert 'city' and 'category' to factors
  mutate(across(c(city, category), as.factor))

zipcodes_data <- sales.zipcodes %>%
  left_join(acs.zipcodes, by = "zipcode") %>%
  # Convert 'zipcode' and 'category' to factors
  mutate(across(c(zipcode, category), as.factor))
```


```{r}
# Define function to check for missing values
missing_check <- function(data){
  missing_count = data %>%
    summarise(across(everything(), ~ sum(is.na(.))))
  print(missing_count)
}

# Find count of missing values
missing_check(counties_data)
missing_check(cities_data)
missing_check(zipcodes_data)
```

```{r}
# Check where is the missings
counties_data %>%
  filter(is.na(high.school)) %>%  
  count(county)   

cities_data %>%
  filter(is.na(high.school)) %>%  
  count(city)

zipcodes_data %>%
  filter(is.na(high.school)) %>%  
  count(zipcode)
```


  - It shows that only County of Obrien data is missing.  

  - Correspondingly, Obrien's all cities (Jewell, Mc Gregor, Saint Ansgar, Saint Charles, and Saint Lucas) data were also missing. 

  - However, three zipcodes of 52004 (Monroe County, Albia),  52036 (Plymouth County, Dike) and 52733 (Muscatine County, Dorchester)
  
  
```{r}
# Remove missing values
counties_data.clean <- na.omit(counties_data)
cities_data.clean <- na.omit(cities_data)
zipcodes_data.clean <- na.omit(zipcodes_data)
```


```{r}
# Define function to generate, income_per_capita, prop_white and prop_black
add_indicator <-function(data){
  data <- data %>%
    mutate(income_per_capita = income/population,
           prop_white = pop.white / population * 100,
           prop_black = pop.black / population * 100)
  return(data)
}

# Add variables into the three dataset
counties_data.clean <- add_indicator(counties_data.clean)
cities_data.clean <- add_indicator(cities_data.clean)
zipcodes_data.clean <- add_indicator(zipcodes_data.clean )
```

```{r}
colnames(counties_data.clean)
```

```{r}
# Define function to get total sale, volume and price in counties, cities and zipcodes level
add_indicator.2  <- function(data, level){
  data %>%
    group_by({{ level }}) %>% 
    mutate(
      total.sale = sum(sale.dollars), 
      total.volume = sum(sale.volume)  
    ) %>%
    select(-c(category, sale.dollars, sale.volume)) %>%  
    distinct()  %>%
    mutate(price = total.sale / total.volume) %>%
    as.data.frame()
}

# For counties, cities and zipcodes data, apply the function
counties_data.clean.2 <- add_indicator.2 (counties_data.clean, county)
cities_data.clean.2 <- add_indicator.2 (cities_data.clean, city)
zipcodes_data.clean.2 <- add_indicator.2 (zipcodes_data.clean, zipcode)
```

  - data.clean contains detailed information for each individual liquor category.
  
  - data.clean.2 aggregates total sales and volume across all categories, providing summarized data for each county, city, and zipcode.
  

# Initial EDA

```{r}
# Define function to get means
mean_stat <- function(data){
  data %>%
    group_by(category) %>%
    summarise(mean_sale = mean(sale.dollars), 
              mean_vloume = mean(sale.volume), 
              median_sale = median(sale.dollars), 
              median_vloume = median(sale.volume), ) 
}

# Descriptive statistic for sales in county level
mean_stat(counties_data.clean) %>%
  kable(digit = 2, 
        caption = "Descriptive statistics for liquor category sales in coutry level")
mean_stat(cities_data.clean)  %>%
  kable(digit = 2, 
        caption = "Descriptive statistics for liquor category sales in city level")
mean_stat(zipcodes_data.clean)  %>%
  kable(digit = 2, 
        caption = "Descriptive statistics for liquor category sales in zipcode level")
```



```{r}
# Summary statistic analysis
desc_fun <- function(data){
  desc_table = data %>%
    select(c(2:6, 14:19)) %>%
    describe() %>%
    select(mean, median, sd, min, max, 
           range, skew, kurtosis)
  return(desc_table)
}

# Descriptive for county data
desc_fun(counties_data.clean.2) %>%
    kable(digits = 2, 
          caption = "Descriptive statistics for aggregated data in coutry level")
  
# Descriptive for cities data
desc_fun(cities_data.clean.2) %>%
    kable(digits = 2, 
          caption = "Descriptive statistics for aggregated data in city level")

# Descriptive for zipcode data
desc_fun(zipcodes_data.clean.2) %>%
    kable(digits = 2, 
          caption = "Descriptive statistics for aggregated data in zipcode level")
```


  - data.clean contains detailed information for each individual liquor category.
  
  - data.clean.2 aggregates total sales and volume across all categories, providing summarized data for each county, city, and zipcode.
  
  
## Relationship between alcohol consumption, education level, unemployment and income


```{r}
# Pairwise scatter plot
pairs.plot <- pairs.panels(
  cities_data.clean.2[
    , c("total.sale", "bachelor",  "unemployment", "income")],
  hist.col = "grey",        
  ellipses = FALSE,        
  lm = TRUE ,             
  cex.cor = 0.6
)
```

  - Alcohol consumption and education level:
  The correlation coefficient between alcohol consumption and education level (bachelor's degree) is 0.25, indicating a positive correlation. The scatter plot also suggests that as the proportion of people with higher education increases, alcohol consumption may increase, although the relationship is not strong.

  - Alcohol consumption and unemployment:
  The correlation coefficient between alcohol consumption and unemployment is 0.07, showing a very weak positive correlation. Scatter plot of alcohol consumption also suggests that unemployment has a positive effect on the level of alcohol consumption but it is very weak.

  - Alcohol consumption and income:
  The scatter plot between alcohol consumption and income indicates a very weak negative correlation (r = -0.05). This implies that income has almost no direct relationship with alcohol consumption, and as income increases, alcohol consumption has a slight downward trend.

  - The distribution of alcohol consumption is severely right-skewed, while bachelor, unemployment, and income are slightly right-skewed


```{r}
# Export data 
write.csv(counties_data.clean, "counties_data.clean.csv")
write.csv(cities_data.clean, "cities_data.clean.csv")
write.csv(zipcodes_data.clean, "zipcodes_data.clean.csv")

write.csv(counties_data.clean.2, "counties_data.clean.2.csv")
write.csv(cities_data.clean.2, "cities_data.clean.2.csv")
write.csv(zipcodes_data.clean.2, "zipcodes_data.clean.2.csv")
```

